\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,mathtools,bm}
\usepackage{booktabs}
\usepackage{hyperref}

\title{MokA: Forward/Backward Math From Code}
\author{Derived directly from repository implementation}
\date{\today}

\begin{document}
\maketitle

\section{Scope and Grounding}
This note is derived from these files:
\begin{itemize}
\item \texttt{models/unified\_arch.py}: multimodal embedding and modality masks.
\item \texttt{models/modeling\_llama.py}: masked LoRA projections used by attention/MLP and LM loss.
\item \texttt{peft\_hyper/tuners/lora.py}: custom \texttt{Linear} with \texttt{lora\_A0/A1/A2} and shared \texttt{lora\_B0}.
\item \texttt{scripts/finetune/finetune.py}: which modules are wrapped and which parameters are trainable.
\item \texttt{trainer.py}: gradient-sensitivity metric definitions.
\end{itemize}

\section{Notation and Shapes}
Per adapted linear layer:
\begin{itemize}
\item Input: $X \in \mathbb{R}^{B\times T\times d_{\text{in}}}$.
\item Base weight: $W_0 \in \mathbb{R}^{d_{\text{out}}\times d_{\text{in}}}$, bias $b$.
\item LoRA branch matrices:
\[
A_0\in\mathbb{R}^{r\times d_{\text{in}}},\quad
A_1\in\mathbb{R}^{r\times d_{\text{in}}},\quad
A_2\in\mathbb{R}^{r\times d_{\text{in}}},\quad
B_0\in\mathbb{R}^{d_{\text{out}}\times r}.
\]
\item Scale $s=\alpha/r$ (code uses \texttt{self.scaling[0]} for all branches).
\item Binary masks from input construction:
\[
M_t,M_v,M_a,M_q\in\{0,1\}^{B\times T\times 1}.
\]
\end{itemize}

\paragraph{Important implementation constraint.}
In code, \texttt{r} is parsed digit-wise (\texttt{str(r)}), and branch outputs are later summed. Therefore branch ranks must match in practice (or the sum would fail).

\section{Multimodal Input Construction}
The model builds one embedding sequence and four aligned masks:
\[
E\in\mathbb{R}^{B\times T\times d_{\text{model}}},\quad
M_t,M_v,M_a,M_q\in\{0,1\}^{B\times T\times 1}.
\]
Text tokens map through token embedding; image/video/audio special tokens are replaced by projector features; labels for non-text inserted modal features are set to $-100$ (ignored in CE). Then sequences are left-padded to common $T$.

\section{Forward: One Adapted Linear Layer}
PyTorch linear convention is
\[
Y_{\text{base}} = XW_0^\top + b.
\]

Define masked inputs:
\[
X_t=X\odot M_t,\quad X_v=X\odot M_v,\quad X_a=X\odot M_a.
\]
With dropout operator $D(\cdot)$:
\[
U_i = s\,A_i(D(X_i)),\quad i\in\{0,1,2\},
\]
where $A_i(\cdot)$ means linear map with weight $A_i$.

\subsection{Cross-modal branch coupling}
For each sample $b$, video branch:
\[
Q_v=U_1^{(b)},\quad K_q=V_q=U_0^{(b)}\odot M_q^{(b)},
\]
\[
S_v=\operatorname{softmax}\!\left(\frac{Q_vK_q^\top}{\sqrt{r}}\right),\quad
O_v=S_vV_q,
\]
\[
\widetilde U_1^{(b)} = U_1^{(b)} + \beta\,(M_v^{(b)}\odot O_v),
\]
with $\beta=\texttt{blc\_weight}$.

Similarly for audio:
\[
\widetilde U_2^{(b)} = U_2^{(b)} + \beta\,(M_a^{(b)}\odot O_a),
\]
where $O_a$ uses query $U_2^{(b)}$ and key/value from question slice of $U_0^{(b)}$.

Then
\[
U_{\Sigma}=U_0+\widetilde U_1+\widetilde U_2,\qquad
Y_{\text{lora}} = B_0(U_{\Sigma}),
\]
and final layer output:
\[
Y = Y_{\text{base}} + Y_{\text{lora}}.
\]

\paragraph{What does ``$B$ multiply $A$'' mean?}
Ignoring masking/cross-attn and using one branch:
\[
Y_{\text{lora},i}=B_0\big(sA_i(X_i)\big)
= X_i(s\,B_0A_i)^\top.
\]
So the effective low-rank update to base weight is
\[
\Delta W_i = s\,B_0A_i,\quad
W_{\text{eff}} = W_0 + \sum_i \Delta W_i.
\]

\section{Where It Is Used}
The wrapped linear layers are injected into
\[
\{q\_proj,k\_proj,v\_proj,o\_proj,gate\_proj,up\_proj,down\_proj\}
\]
across decoder layers. In attention and MLP forward passes, modality masks are passed to these projections.

\section{Language Modeling Loss}
Given logits $Z\in\mathbb{R}^{B\times T\times V}$ and labels $y$:
\[
Z_{\text{shift}}=Z_{:,1:T-1,:},\quad
y_{\text{shift}}=y_{:,2:T}.
\]
Loss is cross-entropy over flattened tokens with ignore index $-100$:
\[
\mathcal L = -\frac{1}{N_{\text{valid}}}\sum_{n\in\mathcal I}
\log\operatorname{softmax}(Z_n)_{y_n}.
\]

\section{Backward Propagation}
\subsection{From CE to logits}
For each valid token row $n$:
\[
\frac{\partial \mathcal L}{\partial Z_n}
=
\frac{1}{N_{\text{valid}}}\left(\operatorname{softmax}(Z_n)-e_{y_n}\right),
\]
and $0$ for ignored rows.

\subsection{Adapter local gradients}
Let $G=\partial\mathcal L/\partial Y_{\text{lora}}$.
For $Y_{\text{lora}}=B_0(U_{\Sigma})$:
\[
\frac{\partial\mathcal L}{\partial B_0}
=
\sum_{b,t} G_{b,t}^{\top}\,U_{\Sigma,b,t},
\qquad
\frac{\partial\mathcal L}{\partial U_{\Sigma}}
=
G\,B_0.
\]

Define split upstreams:
\[
G_0=\frac{\partial\mathcal L}{\partial U_0},\;
G_1=\frac{\partial\mathcal L}{\partial U_1},\;
G_2=\frac{\partial\mathcal L}{\partial U_2},
\]
where $G_1,G_2$ include direct residual path and query-through-attention path.

For branch $i$ with $U_i=s\,A_i(D(X_i))$:
\[
\frac{\partial\mathcal L}{\partial A_i}
=
s\sum_{b,t} G_{i,b,t}^{\top}\,D(X_{i,b,t}).
\]
This gives explicitly:
\[
\boxed{
\frac{\partial\mathcal L}{\partial A_1}
=
s\sum_{b,t} G_{1,b,t}^{\top}\,D(X_{v,b,t})
}
\]
\[
\boxed{
\frac{\partial\mathcal L}{\partial A_2}
=
s\sum_{b,t} G_{2,b,t}^{\top}\,D(X_{a,b,t})
}
\]
with
\[
X_v=X\odot M_v,\quad X_a=X\odot M_a.
\]

\paragraph{What is inside $G_1,G_2$?}
For video (analogous audio):
\[
\widetilde U_1 = U_1 + \beta(M_v\odot O_v),\quad O_v=\text{Attn}(U_1,U_0\odot M_q,U_0\odot M_q).
\]
Hence
\[
\frac{\partial\mathcal L}{\partial U_1}
=
\frac{\partial\mathcal L}{\partial \widetilde U_1}
\left[
I
+
\beta\,\frac{\partial (M_v\odot O_v)}{\partial U_1}
\right].
\]
So $A_1$ receives gradient from both:
\begin{itemize}
\item direct residual term ($I$),
\item attention-query Jacobian term.
\end{itemize}

\subsection{How many gradients are there?}
Per adapted linear module, there are parameter-gradient tensors:
\[
\frac{\partial\mathcal L}{\partial A_0},\;
\frac{\partial\mathcal L}{\partial A_1},\;
\frac{\partial\mathcal L}{\partial A_2},\;
\frac{\partial\mathcal L}{\partial B_0}
\]
(plus any trainable non-LoRA parameters you kept in \texttt{save\_modules}).

If $N_{\text{adapt}}$ modules are wrapped, LoRA matrix-gradient tensor count is:
\[
4N_{\text{adapt}}.
\]
With current target list (7 projections per decoder layer), for $L$ layers:
\[
N_{\text{adapt}} = 7L,\quad \text{LoRA gradient tensors}=28L.
\]

\section{What Gets Updated}
\begin{itemize}
\item Base wrapped weight is frozen in custom \texttt{Linear} (\texttt{weight.requires\_grad=False}).
\item \texttt{mark\_only\_lora\_as\_trainable} first disables non-LoRA params.
\item Then finetune script applies final filter: a parameter is trainable iff its name contains one of \texttt{save\_modules}.
\end{itemize}
So the \emph{actual} update set is:
\[
\Theta_{\text{update}}=\{\theta:\texttt{requires\_grad}(\theta)=\texttt{True after final filter}\}.
\]

\section{Optimizer Update (conceptual)}
For each $\theta\in\Theta_{\text{update}}$, AdamW-style step:
\[
m_t=\beta_1m_{t-1}+(1-\beta_1)g_t,\quad
v_t=\beta_2v_{t-1}+(1-\beta_2)g_t^2,
\]
\[
\hat m_t=\frac{m_t}{1-\beta_1^t},\quad
\hat v_t=\frac{v_t}{1-\beta_2^t},
\]
\[
\theta_{t+1}=\theta_t-\eta\left(\frac{\hat m_t}{\sqrt{\hat v_t}+\epsilon}+\lambda\theta_t\right).
\]

\section{Gradient Sensitivity Metrics in Trainer}
For each group $g\in\{\text{A0,A1,A2,B0,projectors}\}$:
\[
G_g^2 = \sum_{\theta\in g}\left\|\frac{\partial\mathcal L}{\partial\theta}\right\|_F^2,\quad
P_g^2 = \sum_{\theta\in g}\|\theta\|_F^2,
\]
\[
\text{grad\_norm}_g=\sqrt{G_g^2},\quad
\text{param\_norm}_g=\sqrt{P_g^2},\quad
\text{relative\_grad\_norm}_g=\frac{\text{grad\_norm}_g}{\text{param\_norm}_g+\varepsilon}.
\]
These are exactly what hooks/logging in \texttt{trainer.py} compute.

\section{Code-to-Math Trace Checklist}
For any adapted layer in your run:
\begin{enumerate}
\item Build $X,M_t,M_v,M_a,M_q$ from packed multimodal input.
\item Compute base output $XW_0^\top+b$.
\item Compute three masked low-rank branch outputs $U_0,U_1,U_2$.
\item Fuse question-conditioned information into video/audio branches.
\item Sum branches, project by shared $B_0$, add to base.
\item Continue through attention/MLP/stack, then LM head and shifted CE.
\item Backprop: CE $\rightarrow$ logits $\rightarrow$ decoder $\rightarrow$ each adapted linear:
      obtain $\partial\mathcal L/\partial A_0,\partial\mathcal L/\partial A_1,\partial\mathcal L/\partial A_2,\partial\mathcal L/\partial B_0$.
\item Optimizer updates only parameters still marked trainable.
\end{enumerate}

\end{document}
